When Proximity Isn't Enough: Network Segregation and Class Bias in Referrals

### 2025-08-13 writing/manuscript.tex
v1.2.2

\title{When Proximity Isn't Enough: Network Segregation and SES Bias in Referrals}
% \\ \large Experimental Design \\
% \title{Skills, Inequality, and Referrals}
\author{Manuel Munoz\footremember{liser}{Luxembourg Institute of Socio-Economic Research}, Ernesto Reuben\footnote{Division of Social Science, New York University Abu Dhabi} \footrecall{liser}, Reha Tuncer\footnote{University of Luxembourg}}  
% \footnote{Center for Behavioral Institutional Design at New York University Abu Dhabi} 
\begin{document}
\linenumbers 
% HERE FOR LINES
\maketitle

\section*{Abstract}
The share of high-SES connections in one's network is a strong correlate of labor market income. We investigate whether SES biases in referral selection exacerbate differences high-SES connection shares. We conduct a lab-in-the-field experiment with 734 Colombian university students who make incentivized referrals from their enrollment networks. Randomizing participants between performance-only incentives and performance plus a fixed bonus for referral recipients, we find that referrals go to high-performing peers with whom they take many courses together, regardless of incentives. While low-SES referrers exhibit strong in-group preferences, middle- and high-SES referrers show no biases towards their own and other groups, referring along their network shares. We find that network segregation, driven by program selection based on SES, limits cross-SES referral opportunities for even without an explicit SES bias. These suggest institutional policies promoting cross-SES contact are key for reducing SES-based inequalities.
\medskip \\
\textbf{JEL Classification:} C93, J71, D85, Z13 \\
\textbf{Keywords:} inequality, economic mobility, peer networks, class discrimination, homophily

\pagebreak
\section{Introduction}
Equally qualified individuals face different labor market outcomes based on their socioeconomic status \citep{stansbury2024class}. This persistent inequality undermines meritocratic ideals and represents a substantial barrier to economic mobility. A key driver of SES-based inequality in the labor market stems from differences in social capital.\footnote{See for example \citet{loury_dynamic_1977,bourdieu_forms_1986} for pioneering work on the relationship between social position and human capital acquisition.} Economic connectivity, defined as the share of high-SES connections among low-SES individuals, is an important facet of social capital because it correlates strongly with labor market income \citep{chetty2022na}. In this sense, a lack of social capital means lack of access to individuals with influential (higher paid) jobs and job opportunities. It implies having worse outcomes when using one's network to find jobs conditional on the capacity to leverage one's social network.\footnote{See for example \citet{lin1981asr,mouw2003asr} for differential outcomes while using contacts in job search, and \citet{smith2005ajs,Pedulla2019RaceAN} specifically for the effects of race conditional on network use.}

Research on economic connectivity has focused on two distinct mechanisms that shape cross-SES connections: network composition (who you have the chance to meet inside an institutional environment) versus individual preference (who you choose to connect with among those available). The prevailing hypothesis emerging from the seminal work of \citet{chetty2022n} is that increasing exposure to high-SES individuals will lead low-SES individuals to connect with them at higher rates. Universities, in this regard, represent a particularly promising setting as they attract higher-than-population shares of high-SES students, and create more opportunities for cross-SES connections. However, credible evidence on biases in individual preferences to connect across SES groups remains limited. One important reason for this gap is the challenge of creating controlled environments that isolate SES biases while accounting for natural variations in network compositions.

We overcome this challenge through a lab-in-the-field experiment at a Colombian university. Focusing on the role of SES in referral selection, we studied whether individuals who were asked to refer a peer tended to refer a same-SES candidate. We recruited 734 undergraduate students to make incentivized referrals among peers they encountered during their coursework. Referrals were made for the math and critical reading areas of the national university entry exam, and to incentivize performance-based referral selection, participants earned payments up to \$60 per referral based on their nominee's percentile ranking at the university. This setup provided an objective performance benchmark for referrals where SES biases in referral selection could still play a role.

Referrals originated from each participant's unique course enrollment network that we constructed using extensive administrative data. The enrollment network covered each course the referrer had taken with all other undergraduate students at the university (more than 4,500 individuals). It allowed us to observe both characteristics of every potential referral candidate, and the intensity of interaction between the two, which we measured by the number of courses taken together. Referrals from the enrollment networks enabled us to separate network composition (i.e., chance of meeting during coursework and frequency of contact) from SES biases in referral selection (i.e., individual choice in picking a referral). By doing so, we were able to control for naturally varying network compositions with referral candidates at the individual level, and could identify group-level SES biases in referral selection that go beyond mere opportunities to interact at the university.

We randomized participants into two conditions. In the \textbf{Baseline} condition participants made referrals with performance-based incentives only, where their earnings depended on the actual performance of their referrals. In the \textbf{Bonus} condition, participants made referrals with performance-based incentives and an additional fixed bonus (\$25) going to their referral of choice. We designed the \textbf{Bonus} condition to make SES biases in referral selection even more salient. The fixed bonus created incentives to refer peers even if they performed less well, potentially amplifying the relevance of other factors like the SES bias and the connection intensity.

We find that referrals consistently go to higher-performing peers with high connection intensity (14 vs. 4 courses), regardless of the conditions and the exam area. Pooling across these, we find that SES bias in referral selection is primarily driven by low-SES participants exhibiting in-group preferences: Controlling for network composition, low-SES referrers are 45\% more likely to refer other low-SES peers  and 44\% less likely to refer high-SES relative to middle-SES peers. In contrast, middle- and high-SES referrers show no biases toward their own or other groups.

With 93\% of referrals going to peers within the same academic program with whom referrers have taken many courses together, we find that network composition rather than SES biases better explain the observed referral patterns. At the connection intensity where referrals typically occur (median 12 courses together), network segregation becomes stark: low-SES networks contain 44.5\% low-SES peers versus 35\% university-wide (27\% increase), while high-SES networks contain only 15.7\% low-SES peers (55\% decrease from the university average). This segregation means that even without any bias against low-SES peers, high-SES referrers rarely encounter low-SES candidates in their practical choice sets.

@@

Our findings should be interpreted with some scope conditions. First, our referrals have no direct job consequences, and participants refer under anonymity. These may represent a lower stake environment for referrers with no potential reputational concerns. Nevertheless, we replicate typical findings from earlier referral experiments where performance-based incentives brings in qualified candidates from participant networks (e.g., \citet{witte2021workers,beaman_who_2012}).

Second, enrollment networks capture classroom-based interactions and their intensity rather than broader networks of close friendships. While our approach has clear advantages over self-reported friendship network elicitation which suffers from censoring due to limitations in size \citep{griffith2022jle}, triangulating it with an additional method (e.g., social media friendship data) could provide useful for better identifying actual interactions at the university. Still, we find that connection intensity predicts referral selection well beyond same program affiliation, suggesting it does capture meaningful variation in social interactions in some dimension.

@@

Methodologically, we contribute to the literature on job referral experiments. This literature provides causal evidence on why referrals in the labor market are prevalent,\footnote{Referrals solve frictions in the search and matching process and benefit both job-seekers and employers \citep{topa2019social}. Referral candidates tend to get hired more often, have lower turnover, and earn higher wages \citep{obukhova2013ms,brown2016informal,dustmann2016referral}.} finding that performance-based incentives bring in qualified candidates otherwise not identified by demographic characteristics \citep{beaman_who_2012,pallais_why_2016,friebel2023employee,witte2021workers}, and the consequences of relying upon referral hiring, which come at the cost of disadvantaging certain groups \citep{beaman_job_2018,hederos2025gender}. We extend this literature by causally evaluating the effect of a sizeable monetary bonus for the referral candidate and exploring SES biases in referral selection.

@@

\subsection{Inequality and SES in Colombia}
Our experiment took place in Colombia, a country that consistently ranks highly in terms of economic inequality. The richest decile of Colombians earn 50 times more than the poorest decile \citep{barretoherrera2024regional,comisioneconomicaparaamericalatinayel2023}. This economic disparity creates profound differences in outcomes across SES groups in terms of education, geographic residence, language, manners, and social networks \citep{garciavillegas2021dimension, angulo2012movilidad,garcia2015loteria}. While similar patterns also exist elsewhere, Colombia's pronounced economic inequality makes educational and cultural differences across SES groups particularly visible.

In higher education, Colombia's pronounced economic equality manifests itself by preventing meaningful interaction between SES groups. Wealthy families attend exclusive private schools while poorer families access lower-quality public or ``non-elite'' private institutions (see Figure \ref{school_quality_income}). Taken together, the unique ways in which economic inequality manifests itself in the Colombian higher educational setting provides the ideal conditions to study biases related to SES in referral selection.

We rely on Colombia's established estrato classification system to measure SES in our study. In 1994, Colombia introduced a nationwide system that divides the population into six strata based on  ``similar social and economic characteristics'' \citep[p. 102]{hudson_colombia_2010}. Designed for utility subsidies from higher strata to support lower strata, the system aligns with and reinforces existing social class divisions \citep{uribe2008estratificacion, guevara2019spatializing}. It is also widely used by policymakers and in official statistics \citep{fergusson2021desigualdad} and well known to by the public. Using the estrato system, we categorize students in strata 1-2 as low-SES, strata 3-4 as middle-SES, and strata 5-6 as high-SES.

@@

Our study takes place in a non-elite private university which attracts students across the socioeconomic spectrum: The university's undergraduate student body comprises 35\% low-SES, 50\% middle-SES, and 15\% high-SES students.\footnote{Government statistics reveal less than 5\% of the population is high-SES \citep[p. 103]{hudson_colombia_2010}.} This diversity provides opportunities for different SES groups to meet and interact within the same institutional framework.

The opportunities for contact at the university are on equal status. All undergraduate students pay the same fees based on their program choices, and less than 5\% of undergraduate students receive scholarships. The student body is mostly urban ($>70$\%), not part of an ethnic minority ($>95$\%), and has comparable university entry exam scores (see Appendix Figures \ref{math_hist} and \ref{reading_hist}). These make our setting appropriate to study the effects of contact on intergroup discrimination.

@@

Undergraduate students at the university choose among 32 different academic programs. Students take between 5 and 7 courses per semester, and programs last anywhere between 4 and 12 semesters (2 to 6 years). The majority (64\%) of students are enrolled in the 10 programs described in Appendix Figure \ref{program_share}. Medicine, the largest program by size at the university, lasts for 12 semesters, followed by engineering programs at 10 semesters. Most remaining programs last for about 8 to 10 semesters, with specialized programs for immediate entry into the workforce lasting only 4 semesters. Academic program choice thus shapes students' connections at the university, influencing both who they encounter in classes and the frequency of these interactions.

@@

For causal identification, we require two assumptions. First, conditional exogeneity. SES and the number of courses taken together could be endogenous due to program selection. High-SES students sort into expensive programs while low-SES students choose affordable programs, creating SES variation across enrollment networks. Similarly, the number of courses taken together reflects program selection decisions that may correlate with unobserved referral preferences. However, conditional on the realized enrollment network, the remaining variation in both SES and the number of courses taken together across referral candidates must be independent of unobserved factors affecting referral decisions. As a robustness check, we show that being in the same program with the referrer does not impact our SES bias estimates, although it reduces the coefficient estimate for the number of courses taken together.

@@

\section{Design} \label{Design}
We designed an experiment to assess SES biases in referral selection and to evaluate the causal effect of providing a bonus to referral candidates. The experimental design consisted of two incentivized tasks administered in the following sequence: First, participants completed belief elicitation tasks about their own performance on the national university entry exam. Second, they completed the main referral task, nominating peers based on exam performance in two academic areas. Finally, participants reported beliefs about their referrals' performance and provided demographic information. This structure allowed us to measure the accuracy of participants' beliefs and observe their referral decisions in a controlled setting. Figure \ref{timeline} shows the experimental timeline, and detailed instructions are provided in Appendix \ref{instructions}.

\begin{figure}[H]
\centering
\caption{Experimental Timeline}\label{timeline}
\includegraphics[width=\linewidth]{../figures/timeline.jpg}
\begin{tablenotes}
\footnotesize
\item[] \textit{Note:} Participants first reported beliefs about their own university entry exam performance, then made referrals for each academic area, and finally reported beliefs about their referrals' performance and provided demographics.
\end{tablenotes}
\end{figure}

@@

Participants can only refer students with whom they have taken at least one class during their university studies. This requirement ensures that referrals are based on actual peer interactions. We randomize the order in which participants make referrals across the two exam areas.

@@

\subsection{Bonus Treatment}
To examine how different incentive structures affect referral selection, we randomly assign a fixed bonus payment for students who get a referral. In the \textbf{Baseline} treatment, only the participants, i.e., those who make referrals, can earn money based on their referral's performance. The \textbf{Bonus} treatment adds a fixed payment of \$25 to the peer who gets the referral. This payment is independent of the referral's actual performance (see Table \ref{tab:treatments}).

@@

\subsection{Belief elicitation}
We collect two sets of incentivized beliefs to assess the accuracy of participants' knowledge about exam performance. Participants first report beliefs about their own percentile ranking in the university for each exam area. After making referrals, participants report their beliefs about their referrals' percentile ranking in the university. For both belief elicitation tasks, participants earn \$5 per correct belief if their guess is within 7 percentiles of the true value. This margin of error is designed to balance precision with the difficulty of the task.

@@

\section{Sample, Incentives, and Procedure} \label{SID}
We invited all 4,417 undergraduate students who had completed their first year at the university at the time of recruitment to participate in our experiment. A total of 837 students participated in the data collection (19\% response rate). Our final sample consists of 734 individuals who referred peers with whom they had taken at least one class together (88\% success rate).

Table \ref{tab:balance} presents key demographic characteristics and academic performance indicators across treatments (see Appendix Table \ref{tab:selection} for selection). The sample is well-balanced between the \textbf{Baseline} and \textbf{Bonus} conditions, and we observe no statistically significant differences in any of the reported variables (all \textit{p} values $> 0.1$). Our sample is characterized by a majority of middle-SES students with about one-tenth of the sample being high-SES students. The test scores and GPA distributions are balanced. On average, participants had taken 3.8 courses together with members of their network, and the average network consisted of 175 peers.

@@

The experiment was conducted online through Qualtrics, with participants recruited from active students. To ensure data quality while managing costs, we randomly selected one in ten participants for payment. Selected participants received a fixed payment of \$17 for completion. They also received potential earnings from one randomly selected belief question (up to \$5) and one randomly selected referral question (up to \$60). This structure resulted in maximum total earnings of \$82. The average time to complete the survey was 30 minutes, with an average compensation of \$80 for the one in ten participants randomly selected for payment. Payment processing occurred through bank transfer within 15 business days of participation.

@@

\subsection{Network characteristics}
We begin by describing the key features of the enrollment networks. On average, participants connect with 175 other students, and take an average of 3.62 courses together. Figure \ref{connections_class} shows how network characteristics vary by students' time at the university: both the number of connections (network size) and the number of courses taken together (connection intensity) change as participants progress through their studies.

\begin{figure}[H]
  \centering
  \caption{Network size and courses taken together by time spent at the university}\label{connections_class}
  \includegraphics[width=0.6\linewidth]{../figures/connections.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} 
    This figure displays the average number of connections in blue and the average number of courses taken together with connections in gray across semesters completed. Network size (nb. of connections) peaks around 7 semesters before declining as students graduate. Connection intensity (nb. of courses taken) has an increasing trend.
  \end{tablenotes}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \caption{Network characteristics and courses taken together}
  \begin{subfigure}[t]{0.49\textwidth}
    \centering
      \includegraphics[width=\linewidth]{../figures/share_same_program_by_tie.png}
      \caption{Same-program share}  
      \label{share_same_program_by_tie}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \centering
      \includegraphics[width=\linewidth]{../figures/conns_by_tie.png}
      \caption{Network size}
      \label{conns_by_tie}
  \end{subfigure}
  \caption*{\footnotesize\textit{Note:} The left panel illustrates the share of connections within the same program as a function of the number of courses taken together. The right panel shows the average network size as a function of the number of courses taken together. Taking more than 5 courses together with a network member means on average 90\% chance to be in the same program. Similarly, past 5 courses together, the average network size dwindles by 80\%, from more than 210 individuals to below 50. 
  }
\end{figure}

@@

Participants made one referral for math and one referral for the reading part of the university entry exam from their enrollment networks. We observe 1,342 referrals from 734 participants in our final dataset. More than 90\% of these consist of participants referring for both exam areas (see Appendix Table \ref{tab:referral_distribution}). About 70\% of these referrals go to two separate individuals. We compare the outcomes across exam areas for referrals only going to separate individuals in Appendix Table \ref{tab:referral_by_area} and all referrals in Appendix Table \ref{tab:referral_area}. In both cases, we find no meaningful differences between referrals made for math or critical reading areas of the entry exam. As referrals in both exam areas come from the same enrollment network, we group referrals per participant and report average outcomes.

@@

First, referrals go to peers with whom the referrer has taken around 14 courses with on average, compared to almost 4 on average with others in their network (see Figure \ref{tie_hist}). This difference of 10.1 courses is significant ($t = 34.98$, $p < 0.001$), indicating that referrers choose individuals with whom they have higher contact intensity. While the median referral recipient has taken 12 courses together with the referrer, the median network member has shared only 2.8 courses. The interquartile range for referrals spans from 7.5 to 19.5 courses, compared to just 2.1 to 4.0 courses for the broader network, highlighting the concentration of referrals among peers with higher connection intensity. In addition, 93\% of referrals go to students in the same program as the referrer.

@@

Second, we examine entry exam score differences between referred students and the broader network. Referrals go to peers with an average score of 69.5 points, compared to 64.5 points for other network members (see Figure \ref{other_score_hist}). This difference of 5 points is significant ($t = 18.97$, $p < 0.001$), indicating that referrers choose higher-performing peers. While the median referral recipient scores 71 points, the median network member scores 65.1 points. The interquartile range for referrals spans from 65.5 to 75 points, compared to 63.5 to 66.9 points for the broader network, highlighting the  concentration of referrals among higher performing peers.

>>>>

\title{When Proximity Isn't Enough: Network Segregation and Class Bias in Referrals}
% \\ \large Experimental Design \\
% \title{Skills, Inequality, and Referrals}
\author{Manuel Munoz\footremember{liser}{Luxembourg Institute of Socio-Economic Research}, Ernesto Reuben\footnote{Division of Social Science, New York University Abu Dhabi} \footrecall{liser}, Reha Tuncer\footnote{University of Luxembourg}}  
% \footnote{Center for Behavioral Institutional Design at New York University Abu Dhabi} 
\begin{document}
\linenumbers 
% HERE FOR LINES
\maketitle

\section*{Abstract}
The share of high-socioeconomic status (SES) connections in one's network is a strong correlate of labor market income. While universities provide ample opportunities for cross-SES contact, it remains unclear whether this exposure translates into meaningful connections. We investigate this question by exploring SES biases in referral selection. We conduct a lab-in-the-field experiment with 734 Colombian university students who make incentivized referrals from their enrollment networks. Randomizing participants between performance-only incentives and performance plus a fixed bonus for referral recipients, we find that referrals go to high-performing peers with whom they take many courses together, regardless of condition. While low-SES referrers exhibit strong in-group preferences, middle- and high-SES referrers show no biases toward their own and other groups. Network segregation, driven by cost-based program selection, limits cross-SES referral opportunities even without an explicit SES bias. Our results imply that institutional policies promoting cross-SES contact are key for reducing SES-based inequalities.
\medskip \\
\textbf{JEL Classification:} C93, J71, D85, Z13 \\
\textbf{Keywords:} inequality, economic mobility, peer networks, class discrimination, homophily

\section{Introduction}
Equally qualified individuals face different labor market outcomes based on their SES \citep{stansbury2024class}. This persistent inequality undermines meritocratic ideals and represents a substantial barrier to economic mobility. A key driver of SES-based inequality in the labor market stems from differences in social capital.\footnote{See for example \citet{loury_dynamic_1977,bourdieu_forms_1986} for pioneering work on the relationship between social position and human capital acquisition.} Economic connectedness, defined as the share of high-SES connections in one's network, is an important facet of social capital because it correlates strongly with labor market income \citep{chetty2022na}. In this sense, a lack of social capital means lack of access to individuals with influential (higher paid) jobs and related job opportunities. It implies having worse outcomes when using one's network to find jobs conditional on the capacity to leverage one's social network.\footnote{See for example \citet{lin1981asr,mouw2003asr} for differential outcomes while using contacts in job search, and \citet{smith2005ajs,Pedulla2019RaceAN} specifically for the effects of race conditional on network use.}

Research on economic connectedness has focused on two distinct mechanisms that shape cross-SES connections: network composition (who you have the chance to meet inside an institutional environment) versus individual preference (who you choose to connect with among those available). A prevailing hypothesis is that increasing exposure to high-SES individuals will lead to higher rates of cross-SES connections in networks \citep{chetty2022n}. Universities, in this regard, represent a particularly promising setting as they attract higher-than-population shares of high-SES students, and create more opportunities for cross-SES connections. However, whether these cross-SES connection opportunities turn into meaningful contacts, and the role of SES biases in the process has not yet been explored.

We address these questions through a referral experiment in a university setting. Focusing on the role of SES in referral selection, we studied whether individuals tended to refer same-SES peers. We recruited 734 undergraduate students to make incentivized referrals among peers they encountered during their coursework. Referrals were made for the math and critical reading areas of the national university entry exam. To incentivize performance-based referral selection, participants (referrers) earned payments up to \$60 per referral based on their nominee's percentile ranking at the university. This setup provided an objective performance benchmark for referrals where SES biases in referral selection could still play a role.

Referrals originated from each participant's unique course enrollment network that we constructed using extensive administrative data. The enrollment network covered each course that the participant had taken with all other undergraduate students at the university (more than 4,500 individuals). It allowed us to observe every potential referral candidate, and the connection intensity between the candidate and the referrer, which we measured by the number of courses they took together. Referrals from the enrollment networks enabled us to separate network composition (i.e., chance of meeting during coursework and frequency of contact) from SES biases in referral selection (i.e., individual choice in picking a referral). By doing so, we were able to control for naturally varying network compositions with referral candidates at the individual level, and could identify group-level SES biases in referral selection that go beyond mere opportunities to interact at the university.

We randomized participants into two conditions. In the \textbf{Baseline} condition, participants made referrals with performance-based incentives only, where their earnings depended on the actual performance of their referrals. In the \textbf{Bonus} condition, participants made referrals with performance-based incentives and an additional fixed bonus (\$25) going to their referral of choice. The fixed bonus created incentives to refer peers even if they performed less well, potentially amplifying the relevance of other factors like the SES bias and the connection intensity.

We find that referrals consistently go to higher-performing peers with high connection intensity, regardless of conditions and exam areas. Pooling across these, we find that SES bias in referral selection is primarily driven by low-SES participants exhibiting in-group preferences. In our preferred specification, low-SES referrers are 27\% more likely to refer other low-SES peers and 36\% less likely to refer high-SES relative to middle-SES peers. In contrast, middle- and high-SES referrers show no biases toward their own or other groups.

With 93\% of referrals going to peers within the same academic program with whom referrers have taken many courses together, we find that network composition rather than SES biases better explains the observed referral patterns. At the connection intensity where referrals typically occur (median 12 courses together), network segregation becomes stark: low-SES students comprise 44.5\% of low-SES referrers' networks compared to only 15.7\% of high-SES referrers' networks, despite representing 34\% of the university population. This segregation means that even without bias against low-SES peers, high-SES referrers rarely encounter low-SES candidates among their close university connections.

@@

Our findings should be interpreted with some scope conditions. First, our referrals have no direct job consequences, and participants refer under anonymity. These may represent a lower stake environment for referrers. Nevertheless, we replicate typical findings from earlier referral experiments where performance-based incentives brings in qualified candidates from referrer networks (e.g., \citet{witte2021workers,beaman_who_2012}).

Second, enrollment networks capture classroom-based interactions and their intensity rather than broader networks of close friendships. Unlike self-reported friendship networks that must limit the number of connections surveyed \citep{griffith2022jle}, administrative data captures all classroom interactions without artificial size constraints. Combining enrollment networks with  additional network data (e.g., from social media) could be useful for better identifying interactions at the university. Still, we find that connection intensity predicts referral selection well beyond same program affiliation, suggesting it does capture meaningful variation in social interactions.

@@

Third, we contribute to the literature on job referral experiments. This literature provides causal evidence on why referrals in the labor market are prevalent,\footnote{Referrals solve frictions in the search and matching process and benefit both job-seekers and employers \citep{topa2019social}. Referral candidates tend to get hired more often, have lower turnover, and earn higher wages \citep{obukhova2013ms,brown2016informal,dustmann2016referral}.} finding that performance-based incentives bring in qualified candidates otherwise not identified by demographic characteristics \citep{beaman_who_2012,pallais_why_2016,friebel2023employee,witte2021workers}, and the consequences of relying upon referral hiring, which come at the cost of disadvantaging certain groups \citep{beaman_job_2018,hederos2025gender}. We extend this literature by causally evaluating the effects of a sizeable monetary bonus for referral candidates and exploring SES biases in referral selection.

@@

\subsection{Inequality and SES in Colombia}
Our experiment took place in Colombia, a country that consistently ranks highly in terms of economic inequality. The richest decile of Colombians earn 50 times more than the poorest decile \citep{barretoherrera2024regional,comisioneconomicaparaamericalatinayel2023}. This economic disparity creates profound differences in outcomes across SES groups in terms of education, geographic residence, language, manners, and social networks \citep{garciavillegas2021dimension, angulo2012movilidad,garcia2015loteria}.

In higher education, economic inequality manifests itself by preventing meaningful interaction between SES groups. Wealthy families attend exclusive private schools while poorer families access lower-quality public or ``non-elite'' private institutions (see Figure \ref{school_quality_income}). While similar patterns also exist elsewhere, differences in educational outcomes across SES groups are particularly visible in Colombia.

We rely on Colombia's established estrato classification system to measure SES in our study. In 1994, Colombia introduced a nationwide system that divides the population into six strata based on  ``similar social and economic characteristics'' \citep[p. 102]{hudson_colombia_2010}. Designed for utility subsidies from higher strata to support lower strata, the system aligns with and reinforces existing social class divisions \citep{uribe2008estratificacion, guevara2019spatializing}. It is also widely used by policymakers and in official statistics \citep{fergusson2021desigualdad}. Using the estrato system, we categorize students in strata 1-2 as low-SES, strata 3-4 as middle-SES, and strata 5-6 as high-SES.

@@

Our study takes place in a non-elite private university which attracts students across the socioeconomic spectrum: the university's undergraduate student body comprises 34\% low-SES, 51\% middle-SES, and 15\% high-SES students.\footnote{Government statistics reveal less than 5\% of the population is high-SES \citep[p. 103]{hudson_colombia_2010}.} This diversity provides opportunities for different SES groups to meet and interact within the same institutional framework.

The contact at the university is on equal status. All undergraduate students pay the same fees based on their program choices, and less than 5\% of undergraduate students receive scholarships. The student body is mostly urban ($>70$\%), not part of an ethnic minority ($>95$\%), and has similar university entry exam scores (see Appendix Figures \ref{math_hist} and \ref{reading_hist}). These make our setting appropriate to study the effects of contact on intergroup discrimination.

@@

Undergraduate students at the university choose among 32 different academic programs. Students take between 5 and 7 courses per semester, and programs last between 4 and 12 semesters (2 to 6 years). The majority (64\%) of students are enrolled in the 10 programs described in Appendix Figure \ref{program_share}. While medicine, the largest program by size at the university lasts for 12 semesters, specialized programs for immediate entry into the workforce last only 4 semesters. Academic program choice thus shapes students' connections at the university, influencing both who they encounter in classes and the frequency of these interactions.

@@

For identification, we require two assumptions. First, conditional exogeneity. SES and the number of courses taken together could be endogenous due to program selection. High-SES students may sort into expensive programs while low-SES students choose affordable programs, creating SES variation across enrollment networks. Similarly, the number of courses taken together reflects program selection decisions that may correlate with unobserved referral preferences. However, conditional on the realized enrollment network, the remaining variation in both SES and the number of courses taken together across referral candidates must be independent of unobserved factors affecting referral decisions. As a robustness check, we show that being in the same program with the referrer does not impact our SES bias estimates, although it reduces the coefficient estimate for the number of courses taken together.

@@

\section{Design} \label{Design}
We designed an experiment to assess SES biases in referral selection and evaluate the causal effect of providing bonuses to referral candidates. The 30-minute experiment consisted of three sequential tasks: initial belief elicitation about participants' own performance on the national university entry exam, referral tasks where they nominated peers for two exam areas (math and critical reading), and another belief elicitation about their nominees' performance. This structure allowed us to collected incentivized measures for the accuracy of participants' performance beliefs and their referral decisions. Figure \ref{timeline} shows the experiment overview, and detailed instructions are provided in Appendix \ref{instructions}.

\begin{figure}[H]
\centering
\caption{Experiment Overview}\label{timeline}
\includegraphics[width=\linewidth]{../figures/timeline.jpg}
\begin{tablenotes}
\footnotesize
\item[] \textit{Note:} Participants first reported beliefs about their university entry exam performance in two areas, then made referrals for those, and finally reported beliefs about their referrals' performance and provided demographics.
\end{tablenotes}
\end{figure}

@@

Participants are required to only refer students with whom they have taken at least one class during their university studies. Referrals to students without any classes taken together are possible, but not valid. This condition ensures that referrals are based on actual peer interactions. We randomize the order in which participants make referrals across the two exam areas.

@@

\subsection{Bonus Treatment}
To examine how different incentive structures affect referral selection, we randomly assign a fixed bonus payment for students who get a referral (receiver). In the \textbf{Baseline} treatment, only the participants, i.e., those who make referrals (sender), can earn money based on their referral's performance. The \textbf{Bonus} treatment adds a fixed payment of \$25  uniquely to the peer who gets the referral. This payment is independent of the referral's actual performance (see Table \ref{tab:treatments}).

@@

\subsection{Belief elicitation}
We collect two sets of beliefs to assess the accuracy of participants' knowledge about exam performance. Participants first report beliefs about their own percentile ranking in the university for math and critical reading areas. After making referrals, participants report their beliefs about their referrals' percentile ranking in the university. For both belief elicitation tasks, participants earn \$5 per correct belief if their guess is within 7 percentiles of the true value. This margin of error is designed to balance precision with the difficulty of the task.

@@

\section{Sample, Incentives, and Procedure} \label{SID}
We invited all 4,417 undergraduate students who had completed their first semester at the university at the time of recruitment to participate in our experiment. A total of 837 students participated in the data collection (19\% response rate). Our final sample consists of 734 individuals who referred peers with whom they had taken at least one class together, excluding 12\% of participants who made two non-valid referrals.

Table \ref{tab:balance} presents key demographic characteristics and academic performance indicators across treatments (see Appendix Table \ref{tab:selection} for selection). The sample is well-balanced between the \textbf{Baseline} and \textbf{Bonus} conditions, and we observe no statistically significant differences in any of the reported variables (all \textit{p} values $> 0.1$). Our sample is characterized by a majority of middle-SES students with about one-tenth of the sample being high-SES students. The test scores and grade distributions are balanced. On average, participants had taken 3.8 courses together with members of their network, and the average network consisted of 175 peers.

@@

The experiment was conducted online through Qualtrics, and we recruited participants by sending invitations to their student emails. To ensure data quality while managing costs, we randomly selected one in ten participants for payment. Selected participants received a fixed payment of \$17 for completion. They also received potential earnings from one randomly selected belief question (up to \$5) and one randomly selected referral question (up to \$60). This structure resulted in maximum total earnings of \$82. The average time to complete the survey was 30 minutes, with an average compensation of \$80 for one in ten participants randomly selected for payment. Payment processing occurred through bank transfer within 15 business days of participation.

@@

\subsection{Network characteristics}
We begin by describing the key features of the enrollment networks. On average, participants connect with 175 other students, and take an average of 3.8 courses together. Figure \ref{connections_class} shows how network characteristics vary by students' time at the university: both the number of connections (network size) and the number of courses taken together (connection intensity) change as participants progress through their studies.

\begin{figure}[H]
  \centering
  \caption{Network size and courses taken together by time spent at the university}\label{connections_class}
  \includegraphics[width=0.6\linewidth]{../figures/connections.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} 
    This figure displays the average number of connections in gray and the average number of courses taken together with connections in blue across semesters completed. Network size (nb. of connections) peaks around 7 semesters before declining as students graduate. Connection intensity (nb. of courses taken) has an increasing trend.
  \end{tablenotes}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \caption{Network characteristics and courses taken together}
  \begin{subfigure}[t]{0.49\textwidth}
    \centering
      \includegraphics[width=\linewidth]{../figures/share_same_program_by_tie.png}
      \caption{Same-program share}  
      \label{share_same_program_by_tie}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \centering
      \includegraphics[width=\linewidth]{../figures/conns_by_tie.png}
      \caption{Network size}
      \label{conns_by_tie}
  \end{subfigure}
  \caption*{\footnotesize\textit{Note:} Panel (a) illustrates the share of connections within the same program as a function of the number of courses taken together. Panel (b) shows the average network size as a function of the number of courses taken together. Taking more than 5 courses together with a network member means on average 90\% chance to be in the same program. Similarly, past 5 courses together, the average network size dwindles by 80\%, from more than 210 individuals to below 50. 
  }
\end{figure}

@@

Participants made one referral for math and one referral for the reading part of the university entry exam from their enrollment networks. We collected 1,342 valid referrals from 734 participants in our final dataset. More than 90\% of these consist of participants referring for both exam areas (see Appendix Table \ref{tab:referral_distribution}). About 70\% of these referrals go to two separate individuals. We compare the outcomes across exam areas for referrals only going to separate individuals in Appendix Table \ref{tab:referral_by_area} and all referrals in Appendix Table \ref{tab:referral_area}. In both cases, we find no meaningful differences between referrals made for math or critical reading areas of the entry exam. As referrals in both exam areas come from the same enrollment network, we group referrals per participant and report average outcomes.

@@

First, referrals go to peers with whom the referrer has taken around 14 courses with on average, compared to almost 4 on average with others in their network (see Figure \ref{tie_hist}). This difference of 10 courses is significant ($t = 34.98$, $p < 0.001$), indicating that referrers choose individuals with whom they have higher contact intensity. While the median referral recipient has taken 12 courses together with the referrer, the median network member has shared only 2.8 courses. The interquartile range for referrals spans from 7.5 to 19.5 courses, compared to just 2.1 to 4.0 courses for the broader network, highlighting the concentration of referrals among peers with higher connection intensity. In addition, 93\% of referrals go to students in the same program as the referrer.

@@

Second, we examine entry exam score differences between referred students and the broader network. Referrals go to peers with an average score of 69.5 points, compared to 64.5 points for other network members (see Figure \ref{other_score_hist}). This difference of 5 points is significant ($t = 18.97$, $p < 0.001$), indicating that referrers choose higher-performing peers. While the median referral recipient scores 71 points, the median network member scores 65.1 points. The interquartile range for referrals spans from 65.5 to 75 points, compared to 63.5 to 66.9 points for the broader network, highlighting the  concentration of referrals among higher performing peers. Participant beliefs regarding their referral's and own performance rank at the university also support these findings (see Appendix Figures \ref{own_belief} and \ref{other_belief}).

\begin{figure}[H]
  \centering
  \caption{Entry exam scores of network members and referrals}\label{other_score_hist}
  \includegraphics[width=0.6\linewidth]{../figures/other_score_hist.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure compares the distributions of entry exam scores (math and critical reading average) between referrers' network members (orange) versus their chosen referral recipients (dark blue) for all 734 participants. 75\% of referral recipients score above 65.5 points compared to only 25\% of network members scoring above 66.9 points. The distributions are significantly different (Kolmogorov-Smirnov test $D = 71.16$, $p < 0.001$). 
  \end{tablenotes}
\end{figure}
